\vspace*{-2cm}{\bf Authors' Response to the Review of EMSE-D-24-00354:\\ ``Understanding Feedback Mechanisms in Machine Learning Jupyter Notebooks''}

### 2025-02-14 report/reviews/1st round/response.tex

\documentclass[11pt,fleqn]{article}

\usepackage{vmargin}
\setpapersize{A4}
\setmarginsrb{2.5cm}{2.5cm}{2.5cm}{2.5cm}%
{\baselineskip}{2\baselineskip}{baselineskip}{2\baselineskip}
\setlength{\parindent}{0pt}
\setlength{\parskip}{5pt}

\usepackage{caption}
\usepackage{subcaption}

\usepackage{epsfig}
\usepackage{latexsym}
\usepackage{url}
\usepackage{tgheros}
\renewcommand*\familydefault{\sfdefault}

\newcommand{\eline}{\vspace*{.75\baselineskip}}
\newcommand{\Referee}[1]{\eline \noindent {\bf Reviewer comment #1:} \\}
\newcommand{\Us}{\eline \noindent {\bf Response:}\\}
\newcommand{\TBD}{{\bf To Be Done}}
\newcommand{\newreviewer}[1]{\section*{Reviewer #1}\vspace*{-1.05\baselineskip}}

%\usepackage{tikz}
\usepackage[skins,breakable]{tcolorbox}
\tcbset{textmarker/.style={
%
skin=enhancedmiddle jigsaw,breakable,parbox=false,before skip=-1mm, after skip=0mm,
boxrule=0mm,leftrule=2mm,rightrule=2mm,boxsep=0mm,arc=0mm,outer arc=0mm,
left=2mm,right=2mm,top=2mm,bottom=2mm,toptitle=1mm,bottomtitle=1mm,oversize}}

\newtcolorbox{rcomment}{textmarker,colback=gray!15!white,colframe=gray!80!white}

\newenvironment{revcomment}[1][]
{\Referee{#1}\begin{rcomment}}
{\end{rcomment}}

\usepackage{todo}
\usepackage{hyperref}

%customized
\usepackage{xfrac}
\usepackage{framed}
\newcommand{\highlight}[1]{\begin{framed}%
  \noindent\emph{#1}
\end{framed}}

\expandafter\def\expandafter\quote\expandafter{\quote\itshape}

\graphicspath{{../../}}

\usepackage{xspace}
\newcommand{\ING}{ING\xspace}

% more lenient line breaking (avoids text protruding into margins)
\sloppy

\title{\vspace*{-2cm}{\bf Authors' Response to the Review of
 EMSE-D-24-00354:\\
 ``Understanding Feedback Mechanisms in Machine Learning Jupyter Notebooks''}}

\author{Arumoy Shome, Lu\'{i}s Cruz, Diomidis Spinellis, Arie van Deursen}
\date{}

\begin{document}

\maketitle

\newreviewer{Editor}

\begin{revcomment}

Despite the paper has significant strengths, there are numerous significant weaknesses that go beyond the process and period of a major revision. Specifically, there are serious concerns about the sampling procedure and the missing details about the inter-rater reliability analysis and recommendations to practitioners.

Since this paper interests EMSE, I suggest the authors revise and resubmit it as new. When submitting as new, if the authors ask to be assigned to the same editor, I will do my best to assign it to the same reviewers to speed up the revision process.

\end{revcomment}

\Us We thank the editor and reviewers for their valuable feedback -- it helped us improve the paper in this new submission. In the following, we address point-by-point the concerns raised by the reviewers.

\newreviewer{1}

\begin{revcomment}[1.1]
The paper presents an insightful exploration of feedback mechanisms within the machine learning (ML) development lifecycle, focusing specifically on Jupyter notebooks authored in Python. By conducting an analysis of ~297.8 thousand notebooks, the authors identify three key feedback mechanisms: assertions as a form of explicit feedback; print statements and last cell statements as forms of implicit feedback. The findings underscore a widespread reliance on implicit feedback mechanisms in Jupyter notebooks. Despite their utility for automated validation, assertions are found to be underutilized, highlighting a significant gap in current testing practices.

\end{revcomment}

\Us Thank you for your valuable comments.

\begin{revcomment}[1.2]

To date, Jupyter notebooks are pervasively used in the ML development lifecycle, especially in the data preparation and model training phases. Moreover, this year, Python overtook JavaScript as the most popular programming language on GitHub, underscoring a surge in open-source data science and ML projects on the platform [1]. Hence, investigating how data scientists leverage Python Jupyter notebooks in their daily work is a timely and relevant research endeavor, which could inform the evolution of Jupyter Notebook implementations. Also, understanding the nature of feedback mechanisms used in notebooks paves the way to improve overall testing and QA practices in data science projects. In addition, as the authors suggest, this understanding can be useful for optimizing the transition from notebook-based prototyping to production-grade ML code [2].

The dataset contributed by this study—comprising assertions, print statements, and last-cell statements from notebooks—further enhances its relevance and utility, offering a rich resource for future research.

Concerning the final discussion, I wonder whether - based on their findings - the authors have thought of any recommendations for designers of notebook platforms (e.g., Jupyter Notebook, Jupyter Lab, Google Colaboratory, etc.). Integrating such recommendations could potentially enhance the practical relevance of this study.

\end{revcomment}

\Us Revised Section 5.1 Documentation and Technical Debt to incorporate recommendation for notebook designers and developers.

Based on that, we have updated the manuscript with the following text (cf. Section XX):

\begin{quote}
GitHub is a popular platform to host and collaborate on software projects. It offers extensive tools for project management such as issue tracking, pull requests, code reviews and tooling for continuous integration and delivery. In contrast, Kaggle is focused specifically on data science and machine learning. It provides integrated computational resources, a rich repository of datasets, and a platform to host machine learning competitions. While GitHub supports formal project workflows, Kaggle encourages a more informal approach to data exploration and modeling, making it a preferred choice for data science practitioners looking for hands-on experience with developing machine learning models.  
\end{quote}

\Todo{write response based on text above.}
\Todo{refer to the section where the text was updated (XX).}

\begin{revcomment}[1.3]

To the best of my knowledge, this study is original and it is the first to explore feedback mechanisms in ML Python Jupyter notebooks. Moreover, the work is well-situated within the existing literature, offering a thoughtful background and context for its findings.

\end{revcomment}

\Us 
\Todo{say something nice.}

\begin{revcomment}[1.4]
The methodology employed in this study generally appears robust and well-structured.

To address RQ1, the authors undertake an archival study drawing on two distinct sources: GitHub and Kaggle. The unit of analysis, selection criteria for notebooks, and data collection processes are all clearly delineated. However, I suggest that the authors provide additional context about the source platforms for the benefit of researchers who may not be familiar with them - especially Kaggle. Indeed, GitHub and Kaggle have different purposes and present significantly different environments for notebook authoring and sharing.
\end{revcomment}

\Us Revised Section 3.1 Data Collection to include more information regarding the distinction between Kaggle and Github.

Based on that, we updated the manuscriptin with text below (cf. Section XX):

\begin{quote}
GitHub is a popular platform to host and collaborate on software projects. It offers extensive tools for project management such as issue tracking, pull requests, code reviews and tooling for continuous integration and delivery. In contrast, Kaggle is focused specifically on data science and machine learning. It provides integrated computational resources, a rich repository of datasets, and a platform to host machine learning competitions. While GitHub supports formal project workflows, Kaggle encourages a more informal approach to data exploration and modeling, making it a preferred choice for data science practitioners looking for hands-on experience with developing machine learning models.
\end{quote}

\Todo{expand the first line into a response.}

\begin{revcomment}[1.5]

To answer RQ2 and RQ3, the authors delve in a manual analysis of a sample of notebooks, allocating a budget of 200 hours to this task. However, I have the following concerns regarding this procedure.

-  The manual task constitutes qualitative coding/annotation, which is inherently subjective. Nonetheless, only a single annotator (the first author) has been involved unless they were "unable to determine the purpose of the candidate," in which case it would be "additionally analyzed by the second and third author." However, for tasks of this type, it is usually advised to involve multiple annotators - at least to code a first, pilot subset of examples, thereby validating the annotation strategy through inter-rater reliability metrics.  I recommend that the authors elaborate more on how they addressed potential subjective bias within this annotation process.

\end{revcomment}

\Us

Section 3.3 Case studies was revised to provide additional details regarding how the exploratory case studies were conducted.

Additionally, to validate the feedback mechanism taxonomy created by the first author, the second author independently annotated the case study candidates using the codes developed by the first author. We have an inter-rater agreement score of X (highly agreeable).

\Todo{start with an agreeable statement}
\Todo{add revised text from Section 3.3, the same way we do in response 1.4}

\begin{revcomment}[1.6]

While it's reasonable to set a predefined time budget for qualitative analysis, and the sampling strategy used seems sound, the number of analyzed notebooks appears quite limited. Consequently, several findings are drawn from an exceedingly small number of ``case studies,'' sometimes as few as one. Although the exploratory nature of this study is acknowledged, deriving generalizable conclusions from such isolated instances is challenging. Therefore, it would be advisable for the authors to underscore this limitation more prominently in the Threats to Validity section, and reconsider presenting cases based solely on a single example as definitive findings.
\end{revcomment}

\Us

Revised results section to emphasize the exploratory nature of the study. We also explicitly mentioned that due to the qualitative nature of the study, the results should be interpreted with caution.

Additionally we replaced the word ``Finding'' with ``Observation'' in all summary boxes.

In the Threats to validity section, we remind the reader that the results should not be interpreted as generalizable results.

\Todo{explain why we replace Finding by Observation -- don't expect people to get it immediately if you don't explain why. }

\begin{revcomment}[1.7]
Overall, the paper is well-written and well-organized. The explanations provided are clear. The images and charts are well-crafted, providing valuable visual support to the text and enhancing comprehension.

However, the methodology section lacks specific details about the processes and steps involved in addressing RQ1.
\end{revcomment}

\Us

We have to say that we started from observations and patterns of use of notebooks from prior experience and what you typically see in books, courses and videos. Then to validate, we used X methodology...

“Clever way” to say asserts, prints and last cell statements are common patterns in notebooks

OR

Forget about it...

“ we considerably improved the methodology section...” and include text from other responses to comments

\Todo{work on response based on text above}

\begin{revcomment}[1.8]
Moreover, the authors have summarized each finding using a summary box; while this is potentially useful to improve the paper understandability, the number of these boxes is high and, in a few cases, the summarized text is shorter than the summary itself (see Sect. 4.3.3) or takes up a comparable space (see Sect 4.3.5). The result is a certain degree of repetition, while it remains challenging to keep the big picture.
\end{revcomment}

\Us
We tried our best to reduce the size of the summary boxes, especially where the boxes are the same size of larger than the main text itself. Almost all summary boxes were revised, please see the diff document for the exact changes.

\begin{revcomment}[1.9]

On a minor note, from Fig. 2 it is not clear how many notebooks were left after applying the filtering criteria.
\end{revcomment}

\Us
Figure 2 updated to reflect how many notebooks were left out after the filters.

On second thought, avoid taking action because getting this number is proving to be tricky...

\begin{revcomment}[1.10]

The methodology adopted in this study is thoroughly described. Moreover, the authors have released the dataset and replication package of this paper on Figshare, under a CC-BY license. Overall, I believe the study should be quite easy to replicate.

An additional appendix has been shared by the authors, containing the most representative examples referenced in the article. However, this appendix is hosted online on a personal website, rather than being incorporated directly at the end of the paper or included with the replication package. Given that references to this appendix are distributed throughout the paper and considering that publications are ideally self-contained, incorporating the appendix within the paper itself would be preferable. If the appendix is too extensive to be included in full, I recommend publishing it as part of the replication package on Figshare, where its long-term accessibility and persistency can be ensured.
\end{revcomment}

\Us Online appendix added to the replication package as a PDF document.

\Todo{make answer more complete}

\begin{revcomment}[1.11]

\textbf{Minor issues}

\begin{itemize}

  \item - [page 4, lines 44-46] "Jupyter notebooks […] leverage the IPython kernel to execute Python code and extends its capabilities […]" => "extend"

  \item - [page 5, Fig. 1] "Content of the cell are in the source key" => "The content of the cell is in the source key" as well as "The raw content of outputs that produce text are under the text key." => "… is …"

  \item - [page 10, lines 31-33] "Methods such as […] were frequently used to accomodate for the stochastic nature of ML algorithms and comparing high-dimensional data structures." => "for comparing" or "to compare"

  \item - [page 13, lines 33-34] "Figure 16 show" => "Figure 16 shows"

  \item - [page 26, lines 35-36] "Catching data errors are critical" => "Catching data errors is critical"

  \item - Section 5.3: in addition to the course by Kästner and Kang, other courses covering MLOps and Software Engineering for ML are currently fostering ML testing literacy among data science students. Examples are the course on "Software Engineering for AI-enables Systems" [2, 3] held at the University of Bari (Italy) and at the Universitat Politècnica de Catalunya (Spain), as well as the course "Machine Learning Systems Design" held at ULiege (Belgium) [4].

\end{itemize}
\end{revcomment}

\Us We thank the reviewer for flagging all these minor issues. We have addressed all of them with the new version of the manuscript and hope our improvements meet your expectations.

\Todo{Address all minor issues}

%%%%%%%%%%%%%%%%%%%
\newreviewer{2}

\newreviewer{3}

\todos

\end{document}

### 2025-03-10 report/emse24-response.tex
chore: update EMSE response document

% TODO: make sure to update all the quoted texts after updating the discussion section to mention the observations explicitly
\maketitle

@@

Please note that despite our best efforts, the exerpts from the manuscript that are included in this document may be out dated. This is because both this document and the manuscript are extensive and it was challenging to keep everthing in sync. Our sincere apologies for this inconvenience. We recommend that when in doubt, the reviewers consult the diff document for the most up to date changes that were made in the new submission.

@@

\Us Thank you for your feedback. We have revised section 5.1 Documentation and Technical Debt to incorporate recommendation for notebook users and developers. Here is an exerpt from the manuscript that was added to address this comment.

\begin{quote}
  We recommend that ML practitioners document insights derived from implicit feedback mechanisms, especially for visualizations. Visual cues and trends in visualizations are open to interpretation, meaning that different practitioners may interpret the same plot differently (Heer et al., 2010). By documenting these visual insights, practitioners can ensure that the knowledge encapsulated in the visualizations, along with its interpretation and the subsequent decisions made, is preserved and consistently applied throughout the ML lifecycle. Writing documentation from implicit feedback mechanims and keeping the documentation up-to-date requires significant manual effort. Therein also lies an opportunity for notebook designers and developers to incorporate automated techniques for generating such documentation. We hope that the dataset contributed by this study provides a valuable resource to develop and evaluate such automated tools.
\end{quote}

@@

\Us Thank you for your feedback. We revised Section 3.1 Data Collection to include more information regarding the distinction between Kaggle and Github. Following is an exerpt from the manuscript that was added to address this comment.

@@

Following is the entire text from section 3.2 Exploratory Case Study Analysis. We have include the entire section here since we made several revisions to this part of the manuscript.

@@

  During each case study, the first author categorized each candidate using \emph{inducting coding}. The first author analysed the Python code of the candidate to understand its purpose. Additionally, the entire code cell, the previous cell, next cell and the notebook's purpose was analysed to bring in rich context. For case studies where the first author was unable to determine the purpose of the candidate, it was additionally analysed by the second and third authors. Finally, the entire research team held regular meetings throughout this iterative process to ensure consistency in the categorization of the feedback mechanisms.

@@

\Us Thank you for your feedback. We made several adjustments to the manuscript to address this comment. First, the title for section 3.2 Case Studies was revised to \emph{Exploratory Case Studies} to emphasize the exploratory nature of this study. Second, we replaced the word ``Finding'' with ``Observation'' in all summary boxes to remind the reader that the results presented in this study are not meant to be generalizable to a population since we conducted qualitative research. Please note that the diff document does not highlight the changes made to the summary boxes. We tried several different latex implementations to generate the summary boxes but were unable to get it to show in the diff document. Our sincere apologies for this inconvenience.

@@

% TODO: I don't fully agree with this comment. Revising the title for section 3.2 and changing the word ``Finding'' to ``Observation'' should be sufficient. Since we are using case study analysis, the results are not meant to be generalizable to a population. Furthermore, the criticism regarding only have one candidate in some categories is also invalid. By changing the manuscript, we admit that the fault is on our end which is not the case. How should we handle this?

@@

\Us Thank you for your feedback. We made have revised section 3.1 Data Collection to include more details regarding the steps involved in addressing RQ1. Specifically, we have separated the section into two sub sections. Section 3.1.1 Collection of ML Jupyter Notebooks contains text from the last version of the manuscript and describes the data collection procedure to collect the Jupyter notebooks relevant to this study. We added section 3.1.2 Collection of Feedback Mechanisms to the manuscript which describes the steps involved in extracting the feedback mechanisms from the notebooks in further detail.

@@

  The raw contents of each cell is then converted into an AST. The \texttt{NodeVisitor} class from the \texttt{ast} module is used to traverse the AST and collect all \texttt{Assert} nodes. To collect assertions written using Python modules not included in the standard distribution, we also collected all function \texttt{Call} nodes that contain the string ``assert'' in their name. Similarly, all print statements were collected by traversing the AST to find all function \texttt{Call} nodes with the name ``print''. To collect the last statements, we first removed all cells that did not produce any output and consequently extracted the last top-level node from the AST of the code cell. The raw Python code was regenerated from the ASTs using the \texttt{unparse} method provided by the \texttt{ast} module.

  Prior to conducting the analysis for RQ2 and RQ3, we remove duplicate data points resulting in a data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last cell statements.
\end{quote}

@@

\Us Thank you for your feedback. We tried our best to reduce the size of the summary boxes, especially where the boxes are the same size or larger than the main text itself. Almost all summary boxes were revised. Please note however that the diff document does not highlight the changes made to the summary boxes. We tried several different latex implementations to generate the summary boxes but were unable to get it to show in the diff document. Our sincere apologies for this inconvenience

@@

\Us Thank you for your feedback. We updated Figure 2 to reflect how many notebooks were excluded and the number of notebooks left after the filters. We also updated section 3.1.1 Collection of Jupyter Notebooks to explicitly mention the number of notebooks that were excluded after applying the filters and the number of notebooks in the final dataset.

Following is an exerpt from the revised text of section 3.1.1 Collection of Jupyter Notebooks relevant to this comment.

\begin{quote}
  Since the focus of this study is on the analysis of Python code, [\ldots] 107 thousand notebooks were excluded after applying the above filters, leaving a dataset 190.8 thousand notebooks.
\end{quote}

% TODO: update the figure

@@

\Us Thank you for your feedback. We have added section 4.1 Descriptive Statistics to the manuscript which provides additional information regarding the distribution of the number of cells per notebook. Since the data is really large and contains many outliers, we categorized the notebooks into small, medium and large based on the number of cells they contain. Additionally, we also added figures 4, 5 and 6 that visualize the distibution of the number of cells and feedback mechanisms across the notebook size categories.

@@

  By analyzing whether the findings hold consistently across these clusters, the authors could find patterns specific to notebook complexity. This approach would not only validate the robustness of the results but also provide more targeted recommendations for practitioners working with notebooks of varying complexity. Including this stratified analysis would significantly strengthen the study's contributions.
\end{revcomment}

\Us Thank you for your feedback. We have added section 5.6 Complexity of Notebooks that discusses the existing challenges of defining notebook complexity and its potential value to improve the depth of this study as future work. While we agree that analyzing the complexity of notebooks may lead to interesting patterns, it is however difficult to define notebook complexity. We agree that the number of cells (or size) of a notebook is a valid albeit not the only metric that should be considered to define notebook complexity.

@@

Please also see our response to comment 1.5 and the diff document for other revisions that were made to section 3.2 Exploratory Case Study Analysis.

\begin{quote}
  For RQ2 and RQ3, we allocated a fixed [\ldots].

  To address the potential bias introduced by this method, we employ stratified random sampling to select candidates for the case study analysis. The subgroups are created based on the aggregated TF-IDF values of the tokens in each candidate. The candidates are then divided into quartiles based on the aggregate value. As a result, all candidates characterized by rare tokens are grouped within a single bin. By randomly selecting candidates from each bin, we ensure that the final population reflects a diverse mix of candidates, mitigating the risk of overrepresenting those that are trivial yet rare.
\end{quote}

@@

\begin{quote}
  The findings of this study align with [\ldots].

  This study further reveals several avenues for future research. We primarily focused on identifying feedback mechanisms used when writing ML code in Jupyter notebooks and subsequently exploring a limited subset of them. Further research is required to investigate these feedback mechanisms in greater depth and systematically organize them into a taxonomy grounded in specific ML validation tasks. It is important to recognize that machine learning workflows vary significantly across different tasks---such as classification, regression, and clustering. Each task necessitates distinct feedback mechanisms. For instance, assertions regarding image dimensions are pertinent in computer vision, whereas data distribution checks are critical in natural language processing. By analyzing the dataset according to the specific types of ML tasks, researchers can obtain targeted insights, thereby enhancing the practicality and applicability of recommendations for practitioners.
\end{quote}

@@

  Jupyter notebooks allow for out-of-order execution, which impacts the effectiveness of feedback mechanisms. A statement such as print("Accuracy:", accuracy) is only useful if the cell defining accuracy has been executed first. The paper partially explore how such dependencies affect the functionality of feedback mechanisms, leaving a gap in the analysis.

@@

% TODO: update this response once the section is finalized
\begin{quote}
  Code written within Jupyter notebooks can reach a convoluted state since the code cells can be executed in any order. Each time a cell is run, it updates the state of the program within the kernel, leading to the potential for discrepancies if cells defining variables or functions are executed after cells that depend on them. Such out-of-order execution may result in a situation where subsequent cells operate on outdated or undefined values, ultimately complicating the workflow and undermining the reproducibility of ML experiments (Pimentel et al., 2019; Wang et al., 2020b).

  Assertions may enhance the reproducibility of ML experiments by ensuring that consistent validation checks are applied, irrespective of the execution order of the notebook cells (Wang et al., 2020b). Conversely, constructs like \texttt{print("Accuracy:", accuracy)} are only effective if the cell defining accuracy has been executed prior to the print statement. In this paper, we have partially examined how these dependencies influence the functionality of feedback mechanisms, highlighting the complexities involved in maintaining reproducibility in scenarios where cell execution order is not strictly linear. Further investigation in this area could provide important insights into optimizing the use of assertions and other feedback mechanisms in Jupyter notebooks, particularly in the context of diverse ML workflows.
\end{quote}

@@

\Us Thank you for your feedback. As per our original definition of feedback mechanisms from section 1 Introduction, all print statements are considered as feedback mechanisms in this study. Regardless, we have revised the definition of feedback mechanisms in the introduction and we hope that this is sufficient to address this comment.

@@

\begin{quote}
  Computational notebooks---specifically Jupyter notebooks---have been ubiquitously adopted by the ML community for developing ML enabled systems (Pimentel et al., 2019; Quaranta et al., 2021; Psallidas et al., 2019; Perkel, 2018). Jupyter notebooks provide an interactive computing environment that allow users to break the development of complex workflows into smaller, more manageable code chunks. Each notebook is composed of a series of code cells, which can be executed independently in a read-eval-print loop (REPL) style of interactive software development. This REPL approach allows users to write a piece of code, run it, and immediately see the results thus facilitating an incremental and iterative style of software development. This modular structure of notebooks provides users the flexibility to experiment with different approaches and algorithms. Users can test hypotheses, debug issues, and make incremental changes with immediate feedback. The ability to visualize data and results inline and document the process in a narrative form further enhances their utility (Kery et al., 2018; Head et al., 2019; Rule et al., 2018; Chattopadhyay et al., 2020).

  Within this environment, feedback mechanisms are essential for improving code quality and diagnosing issues. Feedback mechanims are tools or techniques intentionally integrated into the code by the developer to gain insight into the execution of the program. Importantly, not all forms of feedback quality as feedback mechanisms. For example, a bug in the code may produce feedback indicating an issue, but because the bug was not intentionally inserted, it does not count as a feedback mechanism.

@@

Following is an exerpt from section 5.2 Automated Data Validation added to address this comment.

>>>>

\maketitle

@@

Please note that despite our best efforts, the excerpts from the manuscript that are included in this document may be outdated. This is because both this document and the manuscript are extensive, and it was challenging to keep everything in sync. Our sincere apologies for this inconvenience. We recommend that when in doubt, the reviewers consult the diff document for the most up-to-date changes that were made in the new submission.

@@

\Us Thank you for your feedback. We have revised section 5.1 Documentation and Technical Debt to incorporate recommendation for notebook users and developers. Here is an excerpt from the manuscript that was added to address this comment.

\begin{quote}
  We recommend that ML practitioners document insights derived from implicit feedback mechanisms, especially for visualizations. Visual cues and trends in visualizations are open to interpretation, meaning that different practitioners may interpret the same plot differently (Heer et al., 2010). By documenting these visual insights, practitioners can ensure that the knowledge encapsulated in the visualizations, along with its interpretation and the subsequent decisions made, is preserved and consistently applied throughout the ML lifecycle. Writing documentation from implicit feedback mechanisms and keeping the documentation up-to-date requires significant manual effort. Therein also lies an opportunity for notebook designers and developers to incorporate automated techniques for generating such documentation. We hope that the dataset contributed by this study provides a valuable resource to develop and evaluate such automated tools.
\end{quote}

@@

\Us Thank you for your feedback. We revised Section 3.1 Data Collection to include more information regarding the distinction between Kaggle and Github. Following is an excerpt from the manuscript that was added to address this comment.

@@

Following is the entire text from section 3.2 Exploratory Case Study Analysis. We have included the entire section here since we made several revisions to this part of the manuscript.

@@

  During each case study, the first author categorized each candidate using \emph{inductive coding}. The first author analyzed the Python code of the candidate to understand its purpose. Additionally, the entire code cell, the previous cell, next cell and the notebook's purpose was analyzed to bring in rich context. For case studies where the first author was unable to determine the purpose of the candidate, it was additionally analyzed by the second and third authors. Finally, the entire research team held regular meetings throughout this iterative process to ensure consistency in the categorization of the feedback mechanisms.

@@

\Us Thank you for your feedback. We made several adjustments to the manuscript to address this comment. First, the title for section 3.2 Case Studies was revised to \emph{Exploratory Case Studies} to emphasize the exploratory nature of this study. Second, we replaced the word ``Finding'' with ``Observation'' in all summary boxes to remind the reader that the results presented in this study are not meant to be generalizable to a population since we conducted qualitative research.

Please note that the diff document does not highlight the changes made to the summary boxes. We tried several different latex implementations to generate the summary boxes but were unable to get it to show in the diff document. Our sincere apologies for this inconvenience.

@@

@@

\Us Thank you for your feedback. We revised section 3.1 Data Collection to include more details regarding the steps involved in addressing RQ1. Specifically, we have separated the section into two subsections. Section 3.1.1 Collection of ML Jupyter Notebooks contains text from the last version of the manuscript and describes the data collection procedure to collect the Jupyter notebooks relevant to this study. We added section 3.1.2 Collection of Feedback Mechanisms to the manuscript which describes the steps involved in extracting the feedback mechanisms from the notebooks in further detail.

@@

  The raw contents of each cell were then converted into an AST. The \texttt{NodeVisitor} class from the \texttt{ast} module was used to traverse the AST and collect all \texttt{Assert} nodes. To collect assertions written using Python modules not included in the standard distribution, we also collected all function \texttt{Call} nodes that contain the string ``assert'' in their name. Similarly, all print statements were collected by traversing the AST to find all function \texttt{Call} nodes with the name ``print''. To collect the last statements, we first removed all cells that did not produce any output and consequently extracted the last top-level node from the AST of the code cell. The raw Python code was regenerated from the ASTs using the \texttt{unparse} method provided by the \texttt{ast} module.

  Prior to conducting the analysis for RQ2 and RQ3, we removed duplicate data points resulting in a data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last cell statements.
\end{quote}

@@

\Us Thank you for your feedback. We tried our best to reduce the size of the summary boxes, especially where the boxes are the same size or larger than the main text itself. Almost all summary boxes were revised. Please note however that the diff document does not highlight the changes made to the summary boxes. We tried several latex implementations to generate the summary boxes but were unable to get it to show in the diff document. Our sincere apologies for this inconvenience.

@@

\Us Thank you for your feedback. We updated section 3.1.1 Collection of Jupyter Notebooks to explicitly mention the number of notebooks that were excluded after applying the filters and the number of notebooks in the final dataset.

Following is an excerpt from the revised text of section 3.1.1 Collection of Jupyter Notebooks relevant to this comment.

\begin{quote}
  Since the focus of this study is on the analysis of Python code, [\ldots] 107 thousand notebooks were excluded after applying the above filters, leaving a dataset of 190.8 thousand notebooks.
\end{quote}

@@

\Us Thank you for your feedback. We have added section 4.1 Descriptive Statistics to the manuscript which provides additional information regarding the distribution of the number of cells per notebook. Since the data is really large and contains many outliers, we categorized the notebooks into small, medium and large based on the number of cells they contain. Additionally, we also added figures~\ref{fig:notebook-size-distribution},~\ref{fig:cell-per-notebook-size-distribution} and~\ref{fig:fms-per-notebook-size-distribution} to the manuscript, that visualize the distribution of the number of cells and feedback mechanisms across the notebook size categories.

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{notebook-size-distribution.pdf}
  \caption{Number of notebooks in the small, medium and large notebook size categories.}
  \label{fig:notebook-size-distribution}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{cell-per-notebook-size-distribution.pdf}
  \caption{Distribution of number of cells in all notebook size categories.}
  \label{fig:cell-per-notebook-size-distribution}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{fms-per-notebook-size-distribution.pdf}
  \caption{Distribution of assertions, prints and last cell statements across all notebook size categories.}
  \label{fig:fms-per-notebook-size-distribution}
\end{figure}

@@

  By analyzing whether the findings hold consistently across these clusters, the authors could find patterns specific to notebook complexity. This approach would not only validate the robustness of the results but also provide more targeted recommendations for practitioners working with notebooks of varying complexity. 
  % TODO: we explored this idea but we observed that large notebooks could be (uninteresting) eg. notebook X while smaller... link this with SS (gives similar weightable to all FMs without making assumptions about complexity or other properties)

  Including this stratified analysis would significantly strengthen the study's contributions.
\end{revcomment}

% TODO also add the benefits of using stratified sampling to ensure diversity in population of FMs; "while not perfect, we use SS to increase diversity of our candidates"
% TODO add example of a really large notebook with simple FM and a small one with complex FM
\Us Thank you for your feedback. We have added section 5.6 Complexity of Notebooks that discusses the existing challenges of defining notebook complexity and its potential value to improve the depth of this study as future work. While we agree that analyzing the complexity of notebooks may lead to interesting patterns, it is however difficult to define notebook complexity. We agree that the number of cells (or size) of a notebook is a valid albeit not the only metric that should be considered to define notebook complexity.

@@

\begin{quote}
  The assertions, print statements and last cell statements are first tokenized---special characters and alphanumeric words shorter than two characters are removed. Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively. The term frequency (TF) for the remaining tokens is calculated and then normalized using their inter-document frequency (IDF). This normalization process assigns higher values to tokens that are less frequently encountered, which can lead to overemphasis of candidates that contain rare tokens. Consider for instance, the trivial print statement \texttt{print("hello, world!")}. Since this statement does not commonly appear in the context of machine learning Jupyter notebooks, it will have a high TF-IDF score but adds little substantive value to the analysis.

  To address the potential bias introduced by this method, we employ stratified random sampling to select candidates for the case study analysis. The subgroups are created based on the aggregated TF-IDF values of the tokens in each candidate. The candidates are then divided into quartiles based on the aggregate value. As a result, all candidates characterized by rare tokens are grouped within a single bin. By randomly selecting candidates from each bin, we ensure that the final population reflects a diverse mix of candidates, mitigating the risk of over representing those that are trivial yet rare.
\end{quote}

@@

\begin{quote}
  This study further reveals several avenues for future research. We primarily focused on identifying feedback mechanisms used when writing ML code in Jupyter notebooks and subsequently exploring a limited subset of them. Further research is required to investigate these feedback mechanisms in greater depth and systematically organize them into a taxonomy grounded in specific ML validation tasks. It is important to recognize that machine learning workflows vary significantly across different tasks---such as classification, regression, and clustering. Each task necessitates distinct feedback mechanisms. For instance, assertions regarding image dimensions are pertinent in computer vision, whereas data distribution checks are critical in natural language processing. By analyzing the dataset according to the specific types of ML tasks, researchers can obtain targeted insights, thereby enhancing the practicality and applicability of recommendations for practitioners.
\end{quote}

@@

  Jupyter notebooks allow for out-of-order execution, which impacts the effectiveness of feedback mechanisms. A statement such as print(``Accuracy:'', accuracy) is only useful if the cell defining accuracy has been executed first. The paper partially explore how such dependencies affect the functionality of feedback mechanisms, leaving a gap in the analysis.

@@

\begin{quote}
  Code written within Jupyter notebooks can reach a convoluted state since the code cells can be executed in any order. Each time a cell is run, it updates the state of the program within the kernel, leading to the potential for discrepancies if cells defining variables or functions are executed after cells that depend on them. Such out-of-order execution may result in a situation where subsequent cells operate on outdated or undefined values, ultimately complicating the workflow and undermining the reproducibility of ML experiments (Pimentel et al., 2019; Wang et al., 2020b).

  The assertions presented in Section 4.3 can be used to enhance the reproducibility of ML experiments by ensuring that consistent validation checks are applied, irrespective of the execution order of the notebook cells (Wang et al., 2020b). Conversely, implicit feedback constructs like \texttt{print(``Accuracy:'', accuracy)} are only effective if the cell defining \texttt{accuracy} has been executed prior to the print statement. In this paper, we have partially examined how these dependencies influence the functionality of feedback mechanisms, highlighting the complexities involved in maintaining reproducibility in scenarios where cell execution order is not strictly linear. Further investigation in this area could provide important insights into optimizing the use of assertions and other feedback mechanisms in Jupyter notebooks, particularly in the context of diverse ML workflows.
\end{quote}

@@

\Us Thank you for your feedback. As per our original definition of feedback mechanisms from section 1 Introduction, all print statements are considered as feedback mechanisms in this study.

Regardless, we have revised the definition of feedback mechanisms in the introduction, clearly emphasized the difference between implicit and explicit feedback, and clarified how they differ from testing.

@@

\begin{quote}
  Within this environment, feedback mechanisms are essential for improving code quality and diagnosing issues. \textbf{Feedback mechanisms are tools or techniques intentionally integrated into the code by the developer to gain insight into the execution of the program}. Importantly, not all forms of feedback quality as feedback mechanisms. For example, a bug in the code may produce feedback indicating an issue, but because the bug was not intentionally inserted, it does not count as a feedback mechanism.

  Notebook users can obtain feedback using implicit or explicit mechanisms. \textbf{Implicit feedback mechanisms require manual validation by the user}. For instance, a common prerequisite for all ML models is that the features in the dataset are all numerical. This can be verified with a print statement in a code cell and manually checking the output. This code statement however, will continue to execute even if the data type of a feature changes in subsequent batches of training data.

  In contrast, \textbf{explicit feedback mechanisms document the expected conditions of the code and halt execution if these conditions are not met}. For instance, the implicit validation method mentioned above can be replaced with an explicit assertion: \texttt{assert all(df[col].dtype in ['int64', 'float64'] for col in df.columns), "Not all columns are of numerical type"}. This assert statement will stop the execution if the condition is not satisfied and provide immediate feedback with a clear message explaining the issue.

  It is important to contextualize feedback mechanisms within the broader framework of software testing. Traditional testing methodologies employ formal tests to explicitly identify discrepancies between expected and actual code functionality. \textbf{While all tests serve as feedback mechanisms, not all feedback mechanisms operate within this formal testing paradigm}. For instance, print statements and outputs from the last statement of code cells offer a more informal approach to gaining insights into the execution status and data flow within a Jupyter notebook. While such implicit feedback types can help practitioners diagnose issues within their code, they necessitate manual validation and interpretation, which can increase the potential for oversight and errors.
\end{quote}

@@

Following is an excerpt from section 5.2 Automated Data Validation added to address this comment.
